---
title: "Phase 1 — Who Is Leaving? Exploratory Analysis"
author: "Zohair Khan"
date: "2026-07-07"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
library(tidyverse)
library(patchwork)
library(corrplot)
library(scales)
library(janitor)
theme_set(theme_minimal(base_size = 12))
accent <- c(No = "#8da0cb", Yes = "#e15759")
```

## Purpose

Before any modeling, answer the HR director's first question plainly: **who is leaving, and which groups are leaving fastest?** Everything flagged here becomes a candidate driver for the formal hypothesis tests in Phase 2.

## Data integrity

```{r load}
hr <- read_csv("data/hr-attrition.csv", show_col_types = FALSE)

tibble(
  rows = nrow(hr),
  cols = ncol(hr),
  missing_values = sum(is.na(hr)),
  duplicate_employee_ids = sum(duplicated(hr$EmployeeNumber))
)
```

The dataset is complete: no missing values, one row per employee. Three columns carry no information (`EmployeeCount`, `StandardHours`, `Over18` are constant for every row) and `EmployeeNumber` is an ID, so all four are excluded from analysis. `EmployeeNumber` is kept as the join key for the risk-score outputs in Phase 3.

```{r drop-constants}
constant_cols <- hr %>% summarise(across(everything(), n_distinct)) %>%
  pivot_longer(everything()) %>% filter(value == 1) %>% pull(name)
constant_cols

hr <- hr %>% select(-all_of(constant_cols))
```

## Overall attrition

```{r overall}
overall <- hr %>% count(Attrition) %>% mutate(share = n / sum(n))
overall

attrition_rate <- overall %>% filter(Attrition == "Yes") %>% pull(share)
```

**`r percent(attrition_rate, accuracy = 0.1)` of the workforce left** (`r sum(hr$Attrition == "Yes")` of `r nrow(hr)` employees). At a flat replacement cost of $15K per leaver — the placeholder parameter this project carries through to the dashboard — that is roughly **$`r comma(sum(hr$Attrition == "Yes") * 15000 / 1e6, accuracy = 0.1)`M per year**.

## Which groups are leaving fastest?

A reusable helper: attrition rate by group, with group sizes visible so small-n groups don't get over-read.

```{r rate-helper}
rate_by <- function(df, col) {
  df %>%
    group_by({{ col }}) %>%
    summarise(n = n(), leavers = sum(Attrition == "Yes"),
              attrition_rate = leavers / n, .groups = "drop") %>%
    arrange(desc(attrition_rate))
}

plot_rate <- function(df, col, title) {
  d <- rate_by(df, {{ col }})
  ggplot(d, aes(x = reorder({{ col }}, attrition_rate), y = attrition_rate)) +
    geom_col(fill = "#e15759") +
    geom_hline(yintercept = attrition_rate_overall, linetype = "dashed") +
    geom_text(aes(label = paste0(percent(attrition_rate, 1), " (n=", n, ")")),
              hjust = -0.05, size = 3.2) +
    scale_y_continuous(labels = percent, limits = c(0, .55)) +
    coord_flip() +
    labs(title = title, x = NULL, y = "Attrition rate",
         caption = "Dashed line = company-wide attrition rate")
}
attrition_rate_overall <- attrition_rate
```

### Overtime is the loudest single signal

```{r overtime}
plot_rate(hr, OverTime, "Employees working overtime leave at ~3x the rate of those who don't")
```

### Department and job role

```{r dept-role, fig.height=7}
p1 <- plot_rate(hr, Department, "Attrition by department")
p2 <- plot_rate(hr, JobRole, "Attrition by job role — Sales Reps stand out")
p1 / p2 + plot_layout(heights = c(1, 2))
```

### Travel, marital status, and work-life balance

```{r travel-marital, fig.height=8}
p3 <- plot_rate(hr, BusinessTravel, "Frequent travelers leave more often")
p4 <- plot_rate(hr, MaritalStatus, "Single employees leave more often")
p5 <- plot_rate(hr, WorkLifeBalance, "Attrition by self-rated work-life balance (1=worst)")
p3 / p4 / p5
```

## Numeric drivers

```{r numeric-drivers, fig.height=8}
num_plot <- function(col, title) {
  ggplot(hr, aes(x = Attrition, y = {{ col }}, fill = Attrition)) +
    geom_boxplot(alpha = 0.75, outlier.alpha = 0.3) +
    scale_fill_manual(values = accent, guide = "none") +
    labs(title = title, x = NULL)
}
(num_plot(Age, "Leavers skew younger") +
 num_plot(MonthlyIncome, "Leavers earn less")) /
(num_plot(TotalWorkingYears, "Leavers are earlier-career") +
 num_plot(YearsAtCompany, "Leavers have shorter tenure")) /
(num_plot(DistanceFromHome, "Leavers commute slightly farther") +
 num_plot(YearsSinceLastPromotion, "Time since promotion: no obvious gap"))
```

## Tenure × income: where the risk concentrates

```{r banding}
banded <- hr %>%
  mutate(
    tenure_band = cut(YearsAtCompany, breaks = c(-1, 1, 3, 6, 10, 50),
                      labels = c("0-1 yr", "2-3 yrs", "4-6 yrs", "7-10 yrs", "10+ yrs")),
    income_quartile = factor(ntile(MonthlyIncome, 4),
                             labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)"))
  ) %>%
  group_by(tenure_band, income_quartile) %>%
  summarise(n = n(), attrition_rate = mean(Attrition == "Yes"), .groups = "drop")

ggplot(banded, aes(income_quartile, tenure_band, fill = attrition_rate)) +
  geom_tile(color = "white") +
  geom_text(aes(label = paste0(percent(attrition_rate, 1), "\n(n=", n, ")")), size = 3) +
  scale_fill_gradient(low = "#f7fbff", high = "#e15759", labels = percent) +
  labs(title = "Attrition concentrates in low-income, early-tenure employees",
       x = "Monthly income quartile", y = "Tenure at company", fill = "Attrition")
```

```{r save-key-figure, include=FALSE}
ggsave("outputs/figures/eda-tenure-income-heatmap.png", width = 9, height = 5, dpi = 150)
```

## How the numeric features relate to each other

Several tenure-family variables travel together — this matters later when interpreting regression coefficients (multicollinearity).

```{r corr, fig.height=8, fig.width=9}
num_vars <- hr %>% select(where(is.numeric), -EmployeeNumber)
corrplot(cor(num_vars), method = "color", type = "lower",
         tl.cex = 0.7, tl.col = "black", order = "hclust")
```

## What we take into Phase 2

Candidate drivers, from strongest visual signal to weakest:

1. **OverTime** — the standout categorical gap (~30% vs ~10%).
2. **JobRole / Department** — Sales Representatives (~40%) and lab technicians well above the mean.
3. **MonthlyIncome / JobLevel** — leavers earn substantially less; income and level are nearly the same variable (r ≈ 0.95).
4. **Age / TotalWorkingYears / YearsAtCompany** — an early-career effect; the tenure family is heavily inter-correlated.
5. **MaritalStatus (Single), BusinessTravel (Frequent)** — moderate gaps.
6. **JobSatisfaction / EnvironmentSatisfaction / WorkLifeBalance** — visible but smaller; test formally.
7. **DistanceFromHome** — small; **YearsSinceLastPromotion** — visually near-zero, worth testing to rule out.

Phase 2 puts every one of these through a formal test with multiple-comparison correction and effect sizes, so we can separate real drivers from noise — and from "statistically significant but practically trivial."
