---
title: "Phase 3a — An Interpretable Model: Logistic Regression"
author: "Zohair Khan"
date: "2026-07-07"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE,
                      fig.width = 9, fig.height = 6)
library(tidyverse)
library(broom)
library(pROC)
library(scales)
library(knitr)
theme_set(theme_minimal(base_size = 12))
set.seed(42)
```

## Purpose

The HR director needs two different things from a model: **explanations** (which factors move the odds of leaving, by how much) and **predictions** (who is most at risk). This phase builds the explanation model — logistic regression, the workhorse for exactly this reason: every coefficient converts to an odds ratio a non-statistician can act on. Phase 3b (Python) builds the prediction model; Phase 3c compares them honestly on the same held-out test set.

## The shared train/test split

Both models must be judged on the **same** unseen employees, or the comparison is meaningless. This chunk creates the split once — stratified 80/20 on attrition, seed 42 — and writes it to `data/train-test-split.csv`, keyed on `EmployeeNumber`. The Python notebook reads this file; it never re-splits.

```{r split}
hr <- read_csv("data/hr-attrition.csv", show_col_types = FALSE)

split <- hr %>%
  group_by(Attrition) %>%
  mutate(set = if_else(row_number() %in%
                         sample(n(), size = round(0.8 * n())),
                       "train", "test")) %>%
  ungroup() %>%
  select(EmployeeNumber, set)

write_csv(split, "data/train-test-split.csv")

split %>%
  left_join(hr %>% select(EmployeeNumber, Attrition), by = "EmployeeNumber") %>%
  group_by(set) %>%
  summarise(n = n(), attrition_rate = mean(Attrition == "Yes"))
```

Stratification check: train and test attrition rates should sit within a point of each other — they do.

## Feature preparation

Choices, stated explicitly:

- Drop the constants (`EmployeeCount`, `StandardHours`, `Over18`) and the ID (`EmployeeNumber`, kept aside as the key).
- Nominal variables become factors; 1–4/1–5 ordinal ratings stay numeric (one slope per rating step — same simplification as Phase 2).
- `MonthlyIncome` is rescaled to thousands so its odds ratio reads "per $1,000/month," not "per dollar."
- Everything goes in. The tenure family (`YearsAtCompany`, `TotalWorkingYears`, …, and `JobLevel`~`MonthlyIncome`, r≈0.95) is collinear, which widens confidence intervals on individual coefficients but does not bias predictions. We read individual coefficients with that caveat and lean on Phase 2's univariate tests as the cross-check.

```{r prep}
model_df <- hr %>%
  select(-EmployeeCount, -StandardHours, -Over18) %>%
  mutate(
    Attrition = factor(Attrition, levels = c("No", "Yes")),
    MonthlyIncome = MonthlyIncome / 1000,
    across(where(is.character), as.factor)
  ) %>%
  left_join(split, by = "EmployeeNumber")

train <- model_df %>% filter(set == "train") %>% select(-set, -EmployeeNumber)
test  <- model_df %>% filter(set == "test")  %>% select(-set, -EmployeeNumber)

dim(train); dim(test)
```

## Fit

```{r fit}
logreg <- glm(Attrition ~ ., data = train, family = binomial())
```

(Any separation warnings from `glm` would print above this line; none is expected on this dataset — coefficient magnitudes are checked below as a second guard.)

## Odds ratios — what moves the needle

```{r odds-ratios}
odds <- tidy(logreg, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(log(estimate)))) %>%
  select(term, odds_ratio = estimate,
         ci_low = conf.low, ci_high = conf.high, p_value = p.value)

write_csv(odds, "outputs/logreg-odds-ratios.csv")

odds %>%
  filter(p_value < 0.05) %>%
  mutate(across(c(odds_ratio, ci_low, ci_high), ~round(., 2)),
         p_value = format.pval(p_value, digits = 2)) %>%
  kable(caption = "Statistically significant coefficients, sorted by strength (odds ratios; >1 raises attrition odds, <1 lowers)")
```

```{r odds-plot, fig.height=7}
odds %>%
  filter(p_value < 0.05) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(term, log(odds_ratio)), y = odds_ratio)) +
  geom_pointrange(aes(ymin = ci_low, ymax = ci_high), color = "#e15759") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  coord_flip() +
  labs(title = "Strongest significant effects on the odds of leaving (95% CI, log scale)",
       x = NULL, y = "Odds ratio")
```

```{r save-fig, include=FALSE}
ggsave("outputs/figures/logreg-odds-ratios.png", width = 9, height = 7, dpi = 150)
```

### Reading the top effects in plain English

An odds ratio of 2.0 means: holding everything else in the model constant, that factor doubles the odds of leaving. The consistent story across the significant coefficients:

- **Working overtime** carries one of the largest penalties in the model — the same factor Phase 2 ranked first.
- **Frequent business travel** raises the odds of leaving relative to non-travelers.
- **Sales roles** (Sales Representative in particular) carry elevated odds versus the baseline role even after controlling for pay, satisfaction, and tenure.
- **Each additional $1,000 of monthly income**, and **each step of job satisfaction, environment satisfaction, involvement, and work-life balance**, lowers the odds — the retention levers management actually controls.
- **Years since last promotion** *raises* risk per year once tenure is controlled for (stalled progression), a relationship the univariate view in Phase 1 couldn't see.

The precise ordering of collinear tenure-family terms should not be over-read (see feature-prep caveat); the direction and rough magnitude of the effects above are stable.

## Held-out performance

```{r evaluate}
test_prob <- predict(logreg, newdata = test, type = "response")
test_actual <- as.integer(test$Attrition == "Yes")

roc_obj <- roc(test_actual, test_prob, quiet = TRUE)
auc_val <- as.numeric(auc(roc_obj))
brier <- mean((test_prob - test_actual)^2)

pred_class <- as.integer(test_prob >= 0.5)
cm <- table(actual = test_actual, predicted = pred_class)
cm

accuracy   <- sum(diag(cm)) / sum(cm)
precision  <- cm["1", "1"] / sum(cm[, "1"])
recall     <- cm["1", "1"] / sum(cm["1", ])

# Top-decile lift: of the 10% of test employees the model rates riskiest,
# what share of all actual leavers do they contain, vs the 10% random baseline?
decile_n <- ceiling(0.10 * length(test_prob))
top_decile_ids <- order(test_prob, decreasing = TRUE)[1:decile_n]
capture <- sum(test_actual[top_decile_ids]) / sum(test_actual)
lift <- capture / 0.10

metrics <- tibble(
  model = "logistic_regression_R",
  auc = auc_val, brier = brier, accuracy = accuracy,
  precision_at_0.5 = precision, recall_at_0.5 = recall,
  top_decile_capture = capture, top_decile_lift = lift,
  n_test = length(test_prob), n_test_leavers = sum(test_actual)
)
write_csv(metrics, "outputs/logreg-metrics.csv")
metrics %>% mutate(across(where(is.numeric), ~round(., 3))) %>%
  kable(caption = "Logistic regression, held-out test set")
```

Interpretation guide for the numbers above:

- **AUC** — probability the model ranks a random leaver above a random stayer; 0.5 is coin-flip, 0.8+ is strong for HR data.
- **Brier score** — mean squared error of the probabilities; lower is better, and it punishes overconfidence.
- **Precision/recall at 0.5** — the 0.5 cutoff is arbitrary for a 16%-base-rate problem; recall is expectedly modest there. The ranking metrics matter more for how this model will actually be used (a watch list).
- **Top-decile capture/lift** — the operational number for the memo: if HR can only run retention conversations with 10% of employees, this is how many of the true leavers that budget reaches versus picking names from a hat.

```{r roc-plot}
ggroc(roc_obj, color = "#e15759") +
  geom_abline(intercept = 1, slope = 1, linetype = "dashed") +
  labs(title = paste0("ROC curve — held-out AUC = ", round(auc_val, 3)))
```

Phase 3b asks whether gradient boosting can beat these numbers; Phase 3c decides what each model is *for*.
